---
title: "01-bulkRNA_QC.runfile"
author: "ENPRC Genomics Core"
date:  "`r Sys.Date()`"
output:
  pdf_document:
    df_print: kable
  html_document: default
---

```{r setup, include=FALSE, eval = FALSE}
knitr::opts_chunk$set(warning = TRUE,
                      message = TRUE, 
                      echo=FALSE, 
                      fig.width = 8, 
                      fig.height = 4)

## packages
# analysis
library(DESeq2) # note: in r 4.3.1, if other packages are loading in environment I get error: 'Package ‘GenomeInfoDb’ version 1.38.1 cannot be unloaded'
library(limma)
library(edgeR)
library(msigdbr)
library(fgsea)
library(gencoreBulk)
library(vegan)

# functional
library(tidyverse)
library(readr)
library(readxl)
library(yaml)
library(dplyr)
library(here)
library(tinytex) # required for PDF output
library(rmdformats) # required for HTML output
library(kableExtra)

# plotting
library(ggtext)
library(ggforce)
library(ggpubr)
library(ggplotify)
library(ggh4x)
library(patchwork)
library(plotly)
library(viridis)

### note: I am trying out keeping all of my functions in separate R files. 
###       This does mean they need to be sourced separately, but it makes it a little easier to view and update them 
#### old:
#source(here("R/limmaEdgeR_functions.R")) 
#source(here("R/heatmap_functions.R"))
#source(here("R/QC_functions.R"))

#### semi-in-use
source(here("R/fitVoomLm.R"))

#### In use:
source(here("R/ggplotMDSv2.R"))
source(here("R/checkRLEv2.R"))
source(here("R/createResTable.R"))
source(here("R/getTopDegTbl.R"))

# I personally like to see my month name, even if it isn't ISO compliant
getDate <- function() {
  format(Sys.Date(), "%Y%b%d")
}
```

# Explanation of report workflow

This runfile is where we can run all the code required for the analysis, which will populate our environment with objects used in the format file. We typically run as much of the more heavier computation in the runfile as possible so that it doesn't take long to render the format file, since we may need to render iteratively as we adjust the format, fix typos, write explanatory notes, or add report sections.

# Resources
## Choosing a model design
https://bioconductor.org/packages/release/workflows/vignettes/RNAseq123/inst/doc/designmatrices.html

## Filtering data with EdgeR
https://www.bioconductor.org/packages/devel/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf (page 14)

# Part 0 - Set-up 
*Run set-up steps manually*
*Edit `config/QC_config.yml` before running script*

note: I don't love using the yml files

## 0.1 - Initiate `bulk` object
This object will hold project metadata, sample metadata, counts tables, and model fits
Do this ONCE per project, to avoid overwriting your object!

```{r init bulk object, eval=FALSE}
################################################################################
### Initiate `bulk` object
################################################################################
bulk <- list()
bulk$tables$limma <- list()
bulk$tables$deseq <- list()
bulk$plots$limma <- list()
bulk$plots$deseq <- list()
bulk$config <- list()

################################################################################
### Project data and directories
################################################################################
## Name of project
bulk$config$project <- 'dev'

## Analyst name
bulk$config$author <- 'Rachel Dockman'

## Directory with STAR genome reference 
bulk$config$refDir <- '/yerkes-cifs/runs/Genome_references/macaca_mulatta/t2t/star_index_T2T/'

## Directory with project output from STAR
bulk$config$alignDir <- '/yerkes-cifs/runs/Analysis/2025_Analyses/c25001_AMARR51PAX/c25001_AMARR51PAX_Processing/S2.7.9a_T2T-ncbi/'

## Is the library 'unstranded' or 'stranded'
bulk$config$libType <- 'unstranded'
```

## 0.2 - Experimental config
These are variables that may be called to streamline runfile automation and ideally, require no changes to the template file prior to knitting. I am still working on integrating this info and testing if it is generalizable enough for diverse experimental conditions.

```{r config, eval=FALSE}
################################################################################
### Experimental design info
################################################################################
## Store whether your analysis is longitudinal/repeated
### These logicals will be used for automated format generation
bulk$config$longitudinal <- TRUE

## If longitudinal, set your timepoints
### the goal of this is to ensure plots are in the correct order 
bulk$config$timeVar <- c(
  'tPt_00' = 'Day00',
  'tPt_01' = 'Day01',
  'tPt_02' = 'Day02',
  'tPt_03' = 'Day04',
  'tPt_04' = 'Day07',
  'tPt_05' = 'Day14'
)
#  'tPt_06' = '48hr'
#  'tPt_07' = 'wk3'
#  'tPt_08' = 'exp2'
#  'tPt_09' = '___'
```

## 0.3 - Sample metadata
Personally, I prefer to modify a csv file in Excel to contain the desired columns with proper formatting rather than parsing through the manifest file provided by the investigator.

You must have columns `fileID`, `sampleID`, and `group` to run this template as-is.
For longitudinal/repeated measures, you also need columns `subjectID` and `tPt`

Example:

|    fileID     |  sampleID   |  grp  | subjectID* | tPt*  |  grp.tPt*  | otherAsNeeded |
|---------------|-------------|-------|------------|-------|------------|---------------|
|c25001-s001_...| REu20-Day-0 | mRNA  | REu20      | Day00 | mRNA.Day00 |rhesus/male/etc|

```{r import metadata, eval=FALSE}
meta <- read.csv(here("notes/metadata.csv"))

if(bulk$config$longitudinal==TRUE){
# generate grp.tPt
  meta$grp.tPt <- paste(meta$grp, meta$tPt, sep = ".")
}

# add meta to bulk object  
bulk$md$sampleTable <- meta
```

# Part 1 - Import data

## 1.1 - Read in STAR output and create DGEList object
Requires your metadata to have the correct fileID!

```{r import counts}
## Assemble full path of each read file
bulk$md$sampleTable$readPath <- 
  here(bulk$config$alignDir, bulk$md$sampleTable$fileID, 
       paste0(bulk$md$sampleTable$fileID, "_ReadsPerGene.out.tab"))

## Parse the read files and save mapping stats
counts_tmp <- parseReadPerGeneFiles(
  file.paths = setNames(bulk$md$sampleTable$readPath, 
                        nm = bulk$md$sampleTable$sampleID), 
  library.type = bulk$config$libType)
bulk$md$mapBins <- counts_tmp$map_bins

## Create DGElist 
### The "group" you assign will be used throughout for modeling and plotting
if(bulk$config$longitudinal==TRUE){
  bulk$dge <- DGEList(counts = counts_tmp$read_counts, 
                      samples = bulk$md$sampleTable, 
                      group = bulk$md$sampleTable$grp.tPt)
} else {
  bulk$dge <- DGEList(counts = counts_tmp$read_counts, 
                      samples = bulk$md$sampleTable, 
                      group = bulk$md$sampleTable$grp)
}

## Summarize sample data for report
bulk$tables$sampleSummary <- bulk$dge$samples %>%
    select(fileID, sampleID, subjectID, group) %>%
    dplyr::group_by(group) %>%
    dplyr::summarize(subjectIDs = paste0(subjectID, collapse = ", "),
                     n_subjects = n()) %>%
    as.data.frame()

rm(counts_tmp)

## Compute and normalize library sizes 
bulk$dge$samples$lib.size <- colSums(bulk$dge$counts)
bulk$dge <- normLibSizes(bulk$dge)
bulk$dge$cpm <- cpm(bulk$dge$counts)
bulk$dge$lcpm <- cpm(bulk$dge$counts, log = TRUE)
```

## 1.2 Assign gene symbols from reference gtf as gene IDs
Note: obtaining these names seems only relevant if you are using an Ensembl reference
```{r assign gene symbols}
## STAR output may use Ensembl IDs as gene IDs, in which case, use gtf to assign gene symbols where possible
geneInfoTab <- read_tsv(here(bulk$config$refDir, "geneInfo.tab"),
                        skip = 1,
                        col_names = c("gene_id","gene_symbol","gene_type"),
                        show_col_types = FALSE)

## Remove samples if there are fewer total fragment counts than features
lowcounts <- colSums(bulk$dge$counts) < nrow(geneInfoTab)
stopifnot(!sum(lowcounts)>0)

gene_names <- plyr::mapvalues(rownames(bulk$dge$counts), 
                              from = geneInfoTab$gene_id,
                              to = geneInfoTab$gene_symbol)

## Save original names and unfiltered data
bulk$prefilt_dge <- bulk$dge

## Names are made unique IF you used Ensembl; NCBI reference should already have unique names
rownames(bulk$dge$counts) <- make.names(gene_names, unique=TRUE)
```

## 1.3 - Add design to `bulk` object
By default, this template uses `grp` for cross-sectional studies and `grp.tPt` for longitudinal studies (aka "group" in the DGElist). Default may not work for your project, so modify as needed. 

This initial design is potentially important for gene filtering, so you should probably include most of the important variables. 
You can choose a different design later when you perform analyses so that your contrast selection goes smoothly. 

You should not include subjectID in this section, since it will be blocked later. If you are concerned about variables such as batch/technician etc, those can be added.

More complex design example:
`~ 0 + grp.tPt*sex` (equivalent to `~ 0 + grp.tPt + sex + grp.tPt:sex`)

```{r set design}
bulk$md$design <- stats::model.matrix(~ 0 + group, data = bulk$dge$samples)
```

# 2 - Filter data
Filter based on the grouping factors or treatment factors involved in differential expression testing

## 2.1 - Plot initial mapping stats
```{r mapping_plot, fig.width=10, fig.height=8}
# save mapping stats plot
bulk$plots$qc$pMapBins <- mappingBinsPlot(bulk$md$mapBins) +
  ggtitle("Reads mapped to reference") +
  theme(axis.text.x = element_text(hjust = 1, size = 8),
        plot.title = element_text(face = "bold", size = 20),
        legend.title = element_text(face = "bold", size = 12)) +
  guides(fill = guide_legend(direction = "horizontal",
                             position = "bottom", 
                             title.position = "top")) 

# view plot
bulk$plots$qc$pMapBins
```

## 2.2 - Filter by expression
```{r prep filter count tab}
## If `design = bulk$md$design` includes nuisance variables, replace with `group = {factor of biological interest}`
bulk$md$keep.exprs <- filterByExpr(bulk$prefilt_dge, design = bulk$md$design)
```

### 2.3 - Generate plots for filtering parameters
Note: I chose to use `bulk$prefilt_dge$counts` instead of `bulk$dge$counts` for easier reproducibility when coming back to this script to regenerate plots

```{r make filtering plots}
## gene filtering
bulk$plots$qc$pGeneFilt <- bulk$prefilt_dge$counts %>%
  dplyr::as_tibble(rownames = "gene") %>%
  dplyr::left_join(dplyr::tibble(gene = names(bulk$md$keep.exprs), keep = bulk$md$keep.exprs), 
                   by = c("gene")) %>%
  tidyr::pivot_longer(cols = -c("gene", "keep"), 
                      names_to = "sampleID", 
                      values_to = "counts") %>%
  dplyr::mutate(sampleID = forcats::fct(.data$sampleID, 
                                        levels = levels(bulk$prefilt_dge$samples$sampleID)),
                is_zero = .data$counts == 0) %>%
  dplyr::group_by(.data$sampleID) %>%
  dplyr::summarize(keep_zero = sum(.data$is_zero & .data$keep),
                   keep_nonzero = sum(!.data$is_zero & .data$keep),
                   remove_zero = sum(.data$is_zero & !.data$keep),
                   remove_nonzero = sum(!.data$is_zero & !.data$keep)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_longer(cols = c(starts_with("keep_"), starts_with("remove_")), 
                      names_to = "cat", 
                      values_to = "n_genes") %>%
  dplyr::mutate(cat = forcats::fct(.data$cat, 
                                   levels = rev(c("keep_nonzero", 
                                                  "keep_zero", 
                                                  "remove_nonzero", 
                                                  "remove_zero")))) %>%
  ggplot(aes(x = .data$sampleID, y = .data$n_genes, fill = .data$cat)) +
  theme_bw() +
  geom_bar(stat="identity") +
  scale_fill_manual(name = NULL,
                    values = c("keep_nonzero"="#209964", "keep_zero"="#C4B31C",
                               "remove_nonzero"="#DB890E", "remove_zero"="#C60A19"),
                    breaks = c("keep_nonzero", "keep_zero",
                               "remove_nonzero", "remove_zero"),
                    labels = c("Kept; reads >0", "Kept; reads =0",
                               "Removed; reads >0", "Removed; reads =0")) +
  ylab("Gene counts") + 
  xlab(NULL) +
  ggtitle("Genes filtered by expression level") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),
        plot.title = element_text(face = "bold", size = 20),
        axis.title.y = element_text(face = "bold"),
        legend.text = element_text(size = 12),
        legend.position = "bottom")

  
## library size plot
bulk$plots$qc$pLibSize <- bulk$prefilt_dge$counts %>%
  as_tibble(rownames = "gene") %>%
  dplyr::left_join(tibble(gene = names(bulk$md$keep.exprs), keep = bulk$md$keep.exprs), 
            by = c("gene")) %>%
  tidyr::pivot_longer(cols = -c(gene, keep), names_to = "sampleID", values_to = "counts") %>%
  dplyr::mutate(sampleID = fct(sampleID, levels = levels(bulk$prefilt_dge$samples$sampleID))) %>%
  dplyr::group_by(sampleID, keep) %>%
    dplyr::summarize(lib.size = sum(counts)) %>%
  ungroup() %>%
  ggplot(aes(x = sampleID, y = lib.size, fill = keep)) +
  theme_bw() +
  geom_bar(stat="identity") +
  scale_fill_manual(name = NULL,
                    values = c("TRUE" = "#209964","FALSE" = "#C60A19"),
                    breaks = c("TRUE", "FALSE"),
                    labels = c("Reads kept", "Reads removed")) +
  ylab("Library size") + 
  xlab(NULL) +
  ggtitle("Library size after gene filtering") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),
        plot.title = element_text(face = "bold", size = 20),
        axis.title.y = element_text(face = "bold"),
        legend.text = element_text(size = 12),
        legend.position = "bottom")

## log library size plot
bulk$plots$qc$pLogLibSize <- bulk$plots$qc$pLibSize +
  scale_y_log10() +
  ylab("log10(Library size)") +
  ggtitle("Log library size after gene filtering") 
```

### 2.3.2 - View filtering and library size plots
```{r plot filtering plots, fig.height=5, fig.width=12}
bulk$plots$qc$pGeneFilt
bulk$plots$qc$pLibSize
bulk$plots$qc$pLogLibSize
```

## 2.4 - Check globin expression
As a quality check, this section pulls out globin-related genes to check for contamination
```{r check globin}
globinGenes <- c('HBA', 'HBA1', 'HBA2', 'HBB', 'HBD', 'HBE1', 'HBG1', 'HBG2', 'HBM', 'HBQ1', 'HBZ', 'HBZP',
                 'Hbb-y', 'Hbb-bh1', 'Hbb-b1', 'Hbb-b2', 'Hba-z', 'Hba-a1', 'Hba-a2')
## globin dataframe
bulk$plots$qc$pGlobinBars <- bulk$dge$counts %>%
  as_tibble(rownames = "gene") %>%
  dplyr::mutate(globin = factor(gene %in% globinGenes, 
                                levels = c("TRUE","FALSE"))) %>%
  dplyr::mutate(globin = recode(globin, 
                                "TRUE" = "globin", "FALSE" = "not_globin")) %>%
  pivot_longer(cols = -c(gene, globin), 
               names_to = "sampleID", 
               values_to = "counts") %>%
  mutate(sampleID = fct(sampleID, levels = levels(bulk$dge$samples$sampleID))) %>%
  group_by(sampleID, globin) %>%
  dplyr::summarize(lib.size = sum(counts)) %>%
  ungroup() %>%
  ggplot(aes(x = sampleID, y = lib.size, fill = globin)) +
  geom_bar(stat="identity") +
  ylab("Library size (sum of counts)") + xlab("Sample ID") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6))

## violin plot of percent globin
bulk$plots$qc$pGlobinVln <- bulk$dge$counts %>%
  as_tibble(rownames = "gene") %>%
  dplyr::mutate(globin = factor(gene %in% globinGenes, 
                                levels = c("TRUE","FALSE"))) %>%
  dplyr::mutate(globin = recode(globin, 
                                "TRUE" = "globin", "FALSE" = "not_globin")) %>%
  pivot_longer(cols = -c(gene, globin), 
               names_to = "sampleID", 
               values_to = "counts") %>%
  mutate(sampleID = fct(sampleID, levels = levels(bulk$dge$samples$sampleID))) %>%
  group_by(sampleID, globin) %>%
  dplyr::summarize(lib.size = sum(counts)) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(sampleID),
              names_from = globin, 
              values_from = lib.size) %>%
  mutate(globinPct = globin/(globin+not_globin)) %>% 
  ggplot(aes(x=1,y=globinPct)) +
  geom_violin(fill = NA) + 
  geom_sina() + 
  theme_bw()
```

### 2.4.1 - Plot globin expression
```{r plot globin, fig.height=6, fig.width=16}
bulk$plots$qc$pGlobinBars
bulk$plots$qc$pGlobinVln
```

## 2.5 - Apply the filter
```{r filter dge object}
## Apply the filter
bulk$dge <- bulk$dge[bulk$md$keep.exprs, , keep.lib.sizes=FALSE]

## Recalculate library size, norm factors, cpm and lcpm
bulk$dge$samples$lib.size <- colSums(bulk$dge$counts)
bulk$dge <- normLibSizes(bulk$dge)
bulk$dge$cpm <- cpm(bulk$dge$counts)
bulk$dge$lcpm <- cpm(bulk$dge$counts, log = TRUE)
```

# 3 - Ordination

limma::plotMDS chooses `x_genes` of the most variable genes to generate distance matrices for ordination.
If `x_genes` are chosen based on all samples, then a PCA plot is created (`gene.selection = "common"`)
If `x_genes` are chosen from pairwise comparisons of all samples, then a PCoA plot is created (`gene.selection = "pairwise"`)

## 3.1 - PCA

```{r PCA, fig.width=8, fig.height=8}
## we will be pulling out the first 4 Principal Components (PCs)
dims_list <- list("1v2"=c(1,2), "2v3"=c(2,3), "3v4"=c(3,4))

bulk$plots$qc$limmaPCA <- lapply(dims_list, function(dims) {
  ggplotMDSv2(bulk$dge, 
              mode = "limma",            # "limma" or "deseq", if "deseq" requires VST data
              dims = dims,
              gene.selection = "common", # "common" or "pairwise"
              nDEGs = 20,                # number of DEGs per PC
              show.labels = FALSE,       # show sample labels?
              sampleID = "sampleID", 
              color = "group", 
              shape = "sex",             # limited to ~ 6 shapes
              alpha = 1,
              ellipse = NULL,            # 
              path = NULL                # usually "subjectID" for longitudinal data
              )})

## plot PCAs
lapply(bulk$plots$qc$limmaPCA, function(x){
# default ggplot2 color scheme
  # x$ord + (ggtexttable(x$envfit, rows = NULL))
# better color scheme
  (x$ord + 
      viridis::scale_color_viridis(option = 'turbo', discrete=TRUE)) + 
    (ggtexttable(x$envfit, rows = NULL))
})

## view DEGs associated with the PCs
bulk$plots$qc$limmaPCA$`1v2`$PC1_DEGs
bulk$plots$qc$limmaPCA$`1v2`$PC2_DEGs
bulk$plots$qc$limmaPCA$`3v4`$PC3_DEGs
bulk$plots$qc$limmaPCA$`3v4`$PC4_DEGs
```

## 3.2 - PCoA

```{r PCoA, fig.width=8, fig.height=8}
bulk$plots$qc$limmaPCoA <- lapply(dims_list, function(dims) {
  ggplotMDSv2(bulk$dge, 
              mode = "limma",              # "limma" or "deseq", if "deseq" requires VST data
              dims = dims,
              gene.selection = "pairwise", # "common" or "pairwise"
              nDEGs = 20,                  # number of DEGs per PC
              show.labels = FALSE,         # show sample labels?
              sampleID = "sampleID", 
              color = "group", 
              shape = "sex",               # limited to ~ 6 shapes
              alpha = 1,
              ellipse = NULL,              # 
              path = NULL                  # usually "subjectID" for longitudinal data
              )})

## plot PCoAs
lapply(bulk$plots$qc$limmaPCoA, function(x){
# default ggplot2 color scheme
  # x$ord + (ggtexttable(x$envfit, rows = NULL))
# better color scheme
  (x$ord + 
      viridis::scale_color_viridis(option = 'turbo', discrete=TRUE)) + 
    (ggtexttable(x$envfit, rows = NULL))
})

## view DEGs associated with the PCo's (labeled as PCs for simplicity)
bulk$plots$qc$limmaPCoA$`1v2`$PC1_DEGs
bulk$plots$qc$limmaPCoA$`1v2`$PC2_DEGs
bulk$plots$qc$limmaPCoA$`3v4`$PC3_DEGs
bulk$plots$qc$limmaPCoA$`3v4`$PC4_DEGs
```

# 4 - DESeq2

While the QC steps above were generated with limma/EdgeR, it is a good idea to compare multiple DE analysis tools.
Note: DESeq2 may not handle longitudinal data as well as limma/EdgeR
Luckily, the same models and contrasts can be used in both DE methods

```{r DESeq2}
## prep object
bulk$deseq$dds.data <- DESeqDataSetFromMatrix(countData = bulk$dge$counts,
                                              colData = bulk$dge$samples,
                                              design = bulk$md$design)

## data transformation
#bulk$deseq$rld <- rlog(bulk$deseq$dds.data,blind = FALSE, fitType = "parametric") # slow, not recommended
bulk$deseq$vst <- vst(bulk$deseq$dds.data,blind = FALSE, fitType = "parametric")   # necessary if you want to make PCAs with mode = "deseq"
bulk$deseq$dds.data <- estimateSizeFactors(bulk$deseq$dds.data) # Add scaling factor
bulk$deseq$normCounts <- counts(bulk$deseq$dds.data, normalized = TRUE)

## run DESeq2
bulk$deseq$dds <- DESeq(bulk$deseq$dds.data)
```

## 4.1 - Compare normalization results
```{r RLE, warning=FALSE}
bulk$plots$qc$pLimmaRLE <- checkRLEv2(bulk, DEmethod = "limma")
bulk$plots$qc$pDESeqRLE <- checkRLEv2(bulk, DEmethod = "DESeq2")
bulk$plots$qc$pLimmaRLE
bulk$plots$qc$pDESeqRLE
```

# 5 - Export count table and write bulk object rds

## 5.1 - Create a subdir name and keep it consistent across dirs

```{r prep for export}
## initiate report count (or increase by 1)
rn <- bulk$config$reportNumber
if(is.null(rn)){
  bulk$config$reportNumber <- "01"
} else {
  bulk$config$reportNumber <- sprintf("%02d", as.numeric(bulk$config$reportNumber) + 1)  
  }

## make subdirectories for `outputs` and `rds`
subdir <- paste0("report-", bulk$config$reportNumber)

if (!dir.exists(here("outputs/", subdir))) {
  dir.create(here("outputs/", subdir))}

if (!dir.exists(here("rds/", subdir))) {
  dir.create(here("rds/", subdir))}
```

## 5.2 - Export count table
```{r}
# Filtered count table
bulk$dge$counts %>% 
  as_tibble(rownames = "gene") %>%
  write_csv(., here("outputs/", subdir, paste0("counts-", getDate(), ".csv")))

# Unfiltered count table
bulk$prefilt_dge$counts  %>% 
  as_tibble(rownames = "gene") %>%
  write_csv(., here("outputs/", subdir, paste0("rawCounts-", getDate(), ".csv")))
```

## 5.3 - Write `bulk` rds
```{r}
bulk %>%
  write_rds(., here("rds", subdir, paste0("QC_bulk-", getDate(), ".rds")))
```

# 6 - Render report
## 6.1 - Report name
```{r render}
report_filename <- paste0(bulk$config$project, "-", 
                          bulk$config$reportNumber, "-",
                          format(Sys.Date(), "%Y%b%d"))
```

## 6.2 - HTML
```{r renderHTML}
rmarkdown::render(input = here("scripts/01-bulkRNA_QC.format.Rmd"),
                  output_file = paste0(report_filename, ".html"),
                  output_dir = here('reports'),
                  output_format = "rmdformats::robobook")
```

## 6.3 - PDF
```{r renderPDF}
tinytex::install_tinytex() # If it is already installed, will ask in terminal if you should reinstall
rmarkdown::render(input = here("scripts/01-bulkRNA_QC.format.Rmd"),
                  output_file = paste0(report_filename, ".pdf"),
                  output_dir = here('reports'),
                  output_format = "pdf_document")
```
