---
title: "02-bulkRNA_DE.runfile"
author: "ENPRC Genomics Core"
date:  "`r Sys.Date()`"
output:
  pdf_document:
    df_print: kable
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = TRUE,
                      message = TRUE, 
                      echo=FALSE, 
                      fig.width = 8, 
                      fig.height = 4)
```

# Libraries
```{r}
# analysis
#library(DESeq2)
library(limma)
library(edgeR)
library(msigdbr)
library(fgsea)
#library(gencoreBulk)

# functional
library(tidyverse)
library(readr)
library(readxl)
library(yaml)
library(dplyr)
library(here)
library(tinytex) # required for PDF output
library(rmdformats) # required for HTML output

# plotting
library(ggtext)
library(ggforce)
library(ggpubr)
library(ggplotify)
library(patchwork)
library(plotly)

## custom functions
source(here("R/limmaEdgeR_functions.R")) 
source(here("R/heatmap_functions.R"))
source(here("R/QC_functions.R"))

# I personally like to see my month name, even if it isn't ISO compliant
getDate <- function() {
  format(Sys.Date(), "%Y%b%d")
}
```

# Explanation of report workflow

This runfile is where we can run all the code required for the analysis, which will populate our environment with objects used in the format file. We typically run as much of the more heavier computation in the runfile as possible so that it doesn't take long to render the format file, since we may need to render iteratively as we adjust the format, fix typos, write explanatory notes, or add report sections.

# Resources
## Choosing a model design
https://bioconductor.org/packages/release/workflows/vignettes/RNAseq123/inst/doc/designmatrices.html

## Filtering data with EdgeR
https://www.bioconductor.org/packages/devel/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf (page 14)

# Part 0 - Set-up 
*Run set-up steps manually*

## Starting with object made in script 01-bulkRNA_QC.runfile.Rmd
```{r load data, eval = FALSE}
bulk <- read_rds(here("rds/report-01/QC_bulk-2026May19.rds"))
```

## Custom functions....
```{r}
## I know, I am trying to get rid of functions. But this is a super useful one for automation.
### Store code as functions when the input is something you want to share in the format file
getSourceText <- function(func,a,b) {
  src <- attr(func, "srcref") # pull out function text
  if (is.null(src)) {return("Something went wrong. Ensure R session has `options(keep.source = TRUE)`")}
  raw_text <- as.character(src)
  trim_text <- tail(head(raw_text, -a), -b) # remove lines that encode extra function info 
  formatted_text <- paste(trim_text, collapse = "\n") # convert function text to string
  return(formatted_text)
}

# My group names in the strip text are too long, so I am labeling them with
# a newline
newLineLabel <- function(string) {
  string <- gsub("\\."," \n",string)
  string
}
```

# Part 1 - Prep design and contrasts

## 1.1 - Set FDR

```{r}
fdr <- 0.05
```

## 1.2 - Check your `design` object and decide if you want to keep it

Remember: By default, this template uses `grp` for cross-sectional studies and `group` for longitudinal studies. Whichever is relevant was stored as "group" when the DGElist object was initialized.

Default may not work for your project. Modify as needed.

```{r set design}
bulk$md$design <- stats::model.matrix(~ 0 + group, data = bulk$dge$samples)
```

## 1.3 - Define contrasts

The contrasts you generate can be used for both DESeq2 and limma/EdgeR.
Generally, you want to structure your contrasts as `testVar - controlVar`. I am working on a way to automate building a contrast matrix, but for now, it is manual.

See 'notes/contrastExamples.R' for some complex contrast building examples

```{r define contrast}
## removing metadata column name from contrast matrix
colnames(bulk$md$design) <- gsub("group", "", colnames(bulk$md$design))

# This is a funky workaround I found so the format file doesn't need to be edited to print the contrasts used
## The contrasts are stored within a function, so modify this function like you normally would define contrasts

contrastFunction <- function(x) {
  makeContrasts(
# Within-group-DNA Day0 effects
  DNA_D01vsD00 = DNA.Day01 - DNA.Day00,
  DNA_D02vsD00 = DNA.Day02 - DNA.Day00,
  
# Within-group-mRNA Day0 effects
  mRNA_D01vsD00 = mRNA.Day01 - mRNA.Day00,
  mRNA_D02vsD00 = mRNA.Day02 - mRNA.Day00,
  
# Within-day group effects
  DNA_D01vsmRNA_D01 = DNA.Day01 - mRNA.Day01,
  DNA_D02vsmRNA_D02 = DNA.Day02 - mRNA.Day02,
  
# Change-from-baseline ignoring group
  Overall_D01vsD00 = (DNA.Day01 + mRNA.Day01)/2 - (DNA.Day00 + mRNA.Day00)/2,
  Overall_D02vsD00 = (DNA.Day02 + mRNA.Day02)/2 - (DNA.Day00 + mRNA.Day00)/2,

  levels = bulk$md$design)
}

## eval() then actually starts the function, running makeContrasts() as normal
bulk$md$contr.matrix <- eval(contrastFunction())

## Using the custom function getSourceText, we can print out the contrasts!
cat(getSourceText(contrastFunction,2,2)) # removes two lines from beginning and end

```

# Part 2 - limma/EdgeR

## 2.1 - Run `voomLmFit()` or `glmQLFit()` to get DE results for limma

```{r calculate limma DEGs, warning=FALSE,message=FALSE}
bulk <- fitVoomLm(bulk,
                  design = bulk$md$design,
                  contr.matrix=bulk$md$contr.matrix,
                  block = bulk$dge$samples$subjectID,
                  sample.weights = TRUE)

bulk$degTab$limma <- createResTable(bulk$fit.contr, bulk$md$contr.matrix)
bulk$degTab$limma <- bulk$degTab$limma %>%
    mutate(signif_dir = ifelse(padj >= fdr, "NotSig",
                           ifelse(log2FoldChange < 0, "Down", "Up")) %>%
             factor(., levels = c("NotSig", "Up", "Down")),
           signif = padj < fdr)
```

## 2.2 - Build QC plots of limma-processed data

This chunk includes for each contrast:
- MA-plots (Average Expression vs Log2FoldChange)
- Volcano plots (Log2FoldChange vs -log(padj))
- p-value histogram

```{r build limma QC plots, fig.width = 10, fig.height = 6, warning = FALSE}
# MAplot
## individual plots
for(i in colnames(bulk$md$contr.matrix)){
  bulk$plots$limma$indMAplot[[i]] <- bulk$degTab$limma %>%
    filter(contrast == i) %>%
    {
      ggplot(data = .) +
        geom_point(data = filter(., !signif),
                   aes(x = AveExpr, y = log2FoldChange),
                   color = "black", size = 0.1) +
        geom_point(data = filter(., signif),
                   aes(x = AveExpr, y = log2FoldChange,
                   color = signif_dir, text = gene),
                   size = 0.2) +
        theme_bw() +
        ylim(-max(abs(.$log2FoldChange)),
             max(abs(.$log2FoldChange))) +
        scale_color_manual(
          values = c("black", "tomato", "dodgerblue"),
          labels = c("No significance", "Upregulated","Downregulated"),
          drop = FALSE,
          guide = guide_legend(override.aes = list(size = 3))) +
        labs(color = "Gene Enrichment",
             title = i,
             x = "Average Expression",
             y = "Log2 Fold Change") +
        geom_hline(yintercept=0, linewidth = .5)
    }
}
## combined plots
bulk$plots$limma$pMAplot <- ggarrange(plotlist = bulk$plots$limma$indMAplot,
                                      ncol = 5,
                                      nrow = ceiling(ncol(bulk$md$contr.matrix)/5),
                                      common.legend = TRUE)

# Volcano plots
## individual plots
for(i in colnames(bulk$md$contr.matrix)){
  bulk$plots$limma$indVolcano[[i]] <- bulk$degTab$limma %>%
    filter(contrast == i) %>%
    {
      ggplot(data = .) +
        geom_point(data = filter(., !signif),
                   aes(x = log2FoldChange, y = -log(padj, base = 10)),
                   color = "black", size = 0.1) +
        geom_point(data = filter(., signif),
                   aes(x = log2FoldChange, y = -log(padj, base = 10),
                   color = signif_dir, text = gene),
                   size = 0.2) +
        theme_bw() +
        xlim(-max(abs(.$log2FoldChange)),
             max(abs(.$log2FoldChange))) +
        scale_color_manual(
          values = c("black", "tomato", "dodgerblue"),
          labels = c("No significance", "Upregulated","Downregulated"),
          drop = FALSE,
          guide = guide_legend(override.aes = list(size = 3))) +
        labs(color = "Gene Enrichment",
             title = i,
             x = "Log2 Fold Change",
             y = "Adjusted P-value") +
        geom_vline(xintercept=0, size = .5) +
        geom_hline(yintercept = -log(0.05, base = 10), size = .5,
                   linetype = "dashed", color = "red") +
        scale_y_continuous(labels = function(i) signif(10^-i,1)) +
        annotate("text", x = -10, y=-log(0.05, base = 10),size = 3,
                 vjust = 2, hjust = 0,label = "p = 0.05", color = "red")
      }
}
## combined plots
bulk$plots$limma$pVolcano <- ggarrange(plotlist = bulk$plots$limma$indVolcano,
                                      ncol = 5,
                                      nrow = ceiling(ncol(bulk$md$contr.matrix)/5),
                                      common.legend = TRUE)

# P-value distribution plots
bulk$plots$limma$pPvalue <- bulk$degTab$limma %>%
  dplyr::mutate(!!paste0("pBH < ", fdr) := ifelse(padj < fdr, "TRUE", "FALSE") %>%
                  factor(., levels = c("TRUE", "FALSE"))) %>%
  ggplot(data = ., aes(x = pvalue, fill = .data[[paste0("pBH < ", fdr)]])) +
  geom_histogram(breaks = seq(0,1,0.01)) +
  scale_fill_manual(values = c("tomato", "black"), drop=FALSE) +
  theme_bw() +
  facet_wrap(~contrast, scales = "free_y")

```

### 2.2.1 - View limma QC plots

```{r plot limma QC, fig.width = 12, fig.height = 16}
bulk$plots$limma$pMAplot
bulk$plots$limma$pVolcano
bulk$plots$limma$pPvalue
```

### 2.2.2 - Interactive plots
The plots above can be made interactive, where hovering over a point brings up the gene name. This should not be done with all contrasts, or it freezes your computer. Instead, pull out a few contrasts to explore and ask the investigator if they want any particular contrasts to be made interactive

```{r interactive limma QC plots, fig.width = 8, fig.height = 6, eval = FALSE}
# set to contrasts of interest for interactive plotting
intContrasts <- c("DNA_D01vsD00","mRNA_D01vsD00", "DNA_D01vsmRNA_D01")

for(i in intContrasts){
  print(ggplotly(bulk$plots$limma$indMAplot[[i]], tooltip = "gene"))
}

```

## 2.3 - Summarize limma DEG counts per contrast

```{r limma DEG summary}
## summarize
bulk$tables$limma$nDEGSummary <- bulk$degTab$limma %>%
  group_by(contrast) %>%
    dplyr::rename(l2fc = log2FoldChange, pBH = padj) %>%
    dplyr::summarize(signif_down = sum(pBH < fdr & l2fc < 0),
              nonsignif = sum(pBH >= fdr),
              signif_up = sum(pBH < fdr & l2fc > 0)) %>%
  dplyr::mutate(signif = signif_up + signif_down,
                signif_signs = paste0(signif_up, " ↑ / ", signif_down, " ↓"))

# making text grob object for use with ggarrange
bulk$plots$limma$nDEGtab <- bulk$tables$limma$nDEGSummary %>%
  dplyr::select(contrast, signif_signs) %>%
  ggtexttable(
    .,
    rows = NULL,
    theme = ttheme("light",
                   padding = unit(c(5, 3), "mm"),
                   tbody.style = tbody_style(
                     color = "black", fill = "white",
                     face = "plain", size = 10,
                     linewidth = 1, linecolor = "white",
                     hjust = as.vector(
                       matrix(c(0, .5), ncol = 2, nrow = nrow(.),
                              byrow = TRUE)),
                     x = as.vector(
                       matrix(c(0.05, .5), ncol = 2, nrow = nrow(.),
                              byrow = TRUE))))
    )

## Bar plot
#### experimental, curious as to if the annotation will consistently be in the correct area
bulk$plots$limma$pDEGSummaryBars <- bulk$tables$limma$nDEGSummary %>%
  pivot_longer(
    cols = c(signif_down, signif_up),
    names_to = "sign",
    values_to = "n_signif") %>%
  dplyr::mutate(
    n_signif = ifelse(sign == "signif_down",
                      -n_signif, n_signif)) %>%
  {
    ggplot(.,aes(y = fct_rev(contrast), x = n_signif, fill = sign)) +
      scale_fill_manual(values = c(
        "signif_down" = "dodgerblue",
        'signif_up' = "tomato")) +
      geom_bar(stat="identity") +
      theme_bw() +
      coord_cartesian(xlim = c(min(.$n_signif),max(.$n_signif)),
                      ylim = c(1,nrow(.)/2),
                      clip = "off") +
      theme(plot.margin = unit(c(2, 3, 1, 1), "cm"),
            axis.text = element_text(color = "black")) +
      annotate("text",
               x = quantile(.[which(.$n_signif <0),]$n_signif,.25),
               y = nrow(.)/2+1,
               label = "Downregulated", vjust = 0) +
      annotate("text",
               x = quantile(.[which(.$n_signif >0),]$n_signif,.75),
               y = nrow(.)/2+1,
               label = "Upregulated", vjust = 0) +
      ylab(NULL) +
      xlab("DEGs per contrast (#)") +
      guides(fill = "none") +
      geom_text(data = .,
                aes(label = signif_signs,
                    y = contrast,
                    x = max(n_signif)*1.15,
                    fill = NULL),
                hjust = 0, size = 3)
  }
```

### 2.3.1 - View summary of limma DEGs
```{r view limma summary, fig.width = 6, fig.height = 6}
bulk$plots$limma$pDEGSummaryBars
```

## 2.4 - Heatmaps of DE genes (limma)
```{r make limma DEG heatmaps, warning = FALSE, fig.width = 12, fig.height = 8}
# prepping list for looping
bulk$plots$limma$DEheatmaps <- list()
bulk$tables$limma$scaleGenes <- as.data.frame(t(scale(t(bulk$dge$lcpm))))

# loop to create list of plots for each contrast
for(i in colnames(bulk$md$contr.matrix)){

# defining contrast and pulling out from bulk$md$contr.matrix
  cmSubset <- bulk$md$contr.matrix %>%
    as.data.frame() %>%
    rownames_to_column("group") %>%
    dplyr::select(group, "comparison" = all_of(i))

# subsetting gene table
  gene_tbl <- bulk$degTab$limma %>%
    getTopDegTbl(., contrast_i = i, groupID = "group",
                 arrange_by = "pvalue", direction = "unequal",
                 padj_cutoff = NULL, slice_n = 50,
                 filter_pattern = NULL) %>%
    dplyr::select(gene, padj, l2fc = log2FoldChange) %>%
    dplyr::mutate(color = dplyr::case_when(
      is.na(padj) ~ "grey",
      padj >= fdr ~ "black",
      l2fc > 0 ~ "tomato",
      l2fc < 0 ~ "dodgerblue",
      TRUE ~ NA_character_))

# getting gene order for heatmap
  ord <- hclust(dist(bulk$tables$limma$scaleGenes[c(gene_tbl$gene),]))$order

# combining z-score, sample table, and gene table and prepping factors for plotting
 bulk$plots$limma$DEheatmaps[[i]] <- bulk$tables$limma$scaleGenes[c(gene_tbl$gene),] %>%
    rownames_to_column("gene") %>%
    tidyr::pivot_longer(!gene,
                 names_to = "sampleID",
                 values_to = "zscore")%>%
    dplyr::inner_join(.,bulk$dge$samples,by="sampleID") %>%
    dplyr::inner_join(.,gene_tbl, by="gene") %>%
    dplyr::inner_join(.,cmSubset, by = "group") %>%
    dplyr::arrange(group,subjectID) %>%
    dplyr::mutate(
      group = factor(group, levels = unique(group)),
      subjectID = factor(subjectID, levels = unique(subjectID)),
      gene = factor(gene, levels = unique(gene)[ord])) %>%
   {
     ggplot(data = .,
            aes(x = subjectID, y = gene, fill = zscore)) +
       geom_tile(show.legend = TRUE) +
       scale_fill_gradient2(low="dodgerblue3", mid="white", high="tomato2",
                            limits = c(min(bulk$tables$limma$scaleGenes),
                                       max(bulk$tables$limma$scaleGenes))) +
       facet_grid(~ group,
                  labeller = labeller(group = newLineLabel),
                  scales="free_x", space="fixed") +
       geom_rect(data = subset(., comparison %in% c(1,-1)),
                 linewidth = 1.5, fill = NA, colour = "black",
                 xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
       labs(title = paste0("Top 50 DE Genes in Contrast: ",i),
            x = NULL, y = NULL) +
       theme_minimal(base_size = 10) +
       theme(plot.title = element_text(size = 16, face = "bold"),
             axis.text.x = element_text(angle = 90, hjust = 1, color = "black"),
             axis.text.y = element_text(color = .$color[ord], size = 10),
             strip.text = element_text(size = 10, face = "bold"))
   }
}

```

### 2.4.1 - Plot DE heatmaps (limma)
```{r plot limma DE heatmaps, fig.width = 10, fig.height = 9}
for(i in colnames(bulk$md$contr.matrix)){
    suppressWarnings(print(ggarrange(bulk$plots$limma$DEheatmaps[[i]], legend = "bottom")))
}
```

# Part 3 - DESeq2

## 3.1 - Get DE results from DESeq2 object

```{r get DEGs from DESeq2}
# using same contrasts defined for limma/edgeR

for(i in colnames(bulk$md$contr.matrix)){
  bulk$deseq$contrasts[[i]] <- results(bulk$deseq$dds,
                                       contrast = c(bulk$md$contr.matrix[,i]),
                                       tidy = TRUE) %>%
    dplyr::rename(gene = row) %>%
    dplyr::mutate(contrast = i)
}

bulk$degTab$deseq <- data.table::rbindlist(bulk$deseq$contrasts)
bulk$degTab$deseq[is.na(bulk$degTab$deseq)] <- 1 # since DESeq adds NA for filtered out padj
bulk$degTab$deseq <- bulk$degTab$deseq %>%
  dplyr::mutate(contrast = factor(contrast, levels = colnames(bulk$md$contr.matrix)),
                signif_dir = ifelse(padj >= fdr, "NotSig",
                                    ifelse(log2FoldChange < 0, "Down", "Up")) %>%
                  factor(., levels = c("NotSig", "Up", "Down")),
                signif = padj < fdr)

```

## 3.2 - Build QC plots of DESeq2-processed data

This chunk includes for each contrast:
- MA-plots (Average Expression vs Log2FoldChange)
- Volcano plots (Log2FoldChange vs -log(padj))
- p-value histogram

```{r DESeq2 QC plots, warning = FALSE}
# MAplot
## individual plots
for(i in colnames(bulk$md$contr.matrix)){
  bulk$plots$deseq$indMAplot[[i]] <- bulk$degTab$deseq %>%
    filter(contrast == i) %>%
    {
      ggplot(data = .) +
        geom_point(data = filter(., !signif),
                   aes(x = log(baseMean), y = log2FoldChange), 
                   color = "black", size = 0.1) +
        geom_point(data = filter(., signif), 
                   aes(x = log(baseMean), y = log2FoldChange, 
                       color = signif_dir, text = gene), 
                   size = 0.2) +
        theme_bw() +  
        ylim(-max(abs(.$log2FoldChange)),
             max(abs(.$log2FoldChange))) +
        scale_color_manual(
          values = c("black", "tomato", "dodgerblue"), 
          labels = c("No significance", "Upregulated","Downregulated"),
          drop = FALSE,
          guide = guide_legend(override.aes = list(size = 3))) +
        labs(color = "Gene Enrichment",
             title = i,
             x = "Log baseMean",
             y = "Log2 Fold Change") + 
        geom_hline(yintercept=0, linewidth = .5) 
    }
}
## combined plots
bulk$plots$deseq$pMAplot <- ggarrange(plotlist = bulk$plots$deseq$indMAplot, 
                                      ncol = 5, 
                                      nrow = ceiling(ncol(bulk$md$contr.matrix)/5), 
                                      common.legend = TRUE)

# Volcano plot
## individual plots
for(i in colnames(bulk$md$contr.matrix)){
  bulk$plots$deseq$indVolcano[[i]] <-  bulk$degTab$deseq %>%
    filter(contrast == i) %>%
    {
      ggplot(data = .) +
        geom_point(data = filter(., !signif),
                   aes(x = log2FoldChange, y = -log(padj, base = 10)), 
                   color = "black", size = 0.1) +
        geom_point(data = filter(., signif), 
                   aes(x = log2FoldChange, y = -log(padj, base = 10), 
                       color = signif_dir, text = gene), 
                   size = 0.2) +
        theme_bw() +
        xlim(-max(abs(.$log2FoldChange)), 
             max(abs(.$log2FoldChange))) +
        scale_color_manual(values = c("black", "tomato", "dodgerblue"), 
                           labels = c("No significance", "Upregulated","Downregulated"),
                           drop = FALSE,
                           guide = guide_legend(override.aes = list(size = 3))) +
        labs(color = "Gene Enrichment",
             title = i,
             x = "Log2 Fold Change",
             y = "Adjusted P-value") +
        geom_vline(xintercept=0, size = .5) +
        geom_hline(yintercept = -log(0.05, base = 10), size = .5, 
                   linetype = "dashed", color = "red") +
        scale_y_continuous(labels = function(i) signif(10^-i,1)) +
        annotate("text", x = -10, y=-log(0.05, base = 10),size = 3,
                 vjust = 2, hjust = 0,label = "p = 0.05", color = "red")
    }
}
## combined plots
bulk$plots$deseq$pVolcano <- ggarrange(plotlist = bulk$plots$deseq$indVolcano, 
                                      ncol = 5, 
                                      nrow = ceiling(ncol(bulk$md$contr.matrix)/5), 
                                      common.legend = TRUE)

## P-value distribution plots
bulk$plots$deseq$pPvalue <- bulk$degTab$deseq %>%
      dplyr::mutate(!!paste0("pBH < ", fdr) := ifelse(padj < fdr, "TRUE", "FALSE") %>%
               factor(., levels = c("TRUE", "FALSE"))) %>%
    ggplot(data = ., aes(x = pvalue, fill = .data[[paste0("pBH < ", fdr)]])) +
      geom_histogram(breaks = seq(0,1,0.01)) +
      scale_fill_manual(values = c("tomato", "black"), drop=FALSE) +
      theme_bw() +
    facet_wrap(~contrast, scales = "free_y")
```

### 3.2.1 - View DESeq2 QC plots

```{r plot deseq QC, fig.width = 12, fig.height = 16}
bulk$plots$deseq$pMAplot
bulk$plots$deseq$pVolcano
bulk$plots$deseq$pPvalue
```

### 3.2.2 - Interactive plots
The plots above can be made interactive, where hovering over a point brings up the gene name. This should not be done with all contrasts, or it freezes your computer. Instead, pull out a few contrasts to explore and ask the investigator if they want any particular contrasts to be made interactive

```{r interactive QC plots, fig.width = 8, fig.height = 6}
intContrasts <- c("DNA_D01vsD00","mRNA_D01vsD00", "DNA_D01vsmRNA_D01")

for(i in intContrasts){
  print(ggplotly(bulk$plots$deseq$indMAplot[[i]], tooltip = "gene"))
}
```

## 3.3 - Summarize DESeq2 DEG counts per contrast 

```{r make deseq summary, fig.width = 6, fig.height = 8}
## summarize
bulk$tables$deseq$nDEGSummary <- bulk$degTab$deseq %>%
  dplyr::mutate(padj = replace_na(padj,1)) %>%
  dplyr::rename(l2fc = log2FoldChange, pBH = padj) %>% 
  dplyr::group_by(contrast) %>%
  dplyr::summarize(signif_down = sum(pBH < fdr & l2fc < 0),
                   nonsignif = sum(pBH >= fdr),
                   signif_up = sum(pBH < fdr & l2fc > 0)) %>%
  dplyr::mutate(signif = signif_up + signif_down,
                signif_signs = paste0(signif_up, " ↑ / ", signif_down, " ↓"))

## table
bulk$plots$deseq$nDEGtab <- bulk$tables$deseq$nDEGSummary %>%
  dplyr::select(contrast, signif_signs) %>%
  ggtexttable(
    .,
    rows = NULL,
    theme = ttheme("light",
                   padding = unit(c(5, 3), "mm"),
                   tbody.style = tbody_style(
                     color = "black", fill = "white",
                     face = "plain", size = 10,
                     linewidth = 1, linecolor = "white",
                     hjust = as.vector(
                       matrix(c(0, .5), ncol = 2, nrow = nrow(.), 
                              byrow = TRUE)),
                     x = as.vector(
                       matrix(c(0.05, .5), ncol = 2, nrow = nrow(.), 
                              byrow = TRUE))))
    )

## bar plot
#### experimental, curious as to if the annotation will consistently be in the correct area
bulk$plots$deseq$pDEGSummaryBars <- bulk$tables$deseq$nDEGSummary %>%
  pivot_longer(cols = c(signif_down, signif_up), 
               names_to = "sign", 
               values_to = "n_signif") %>%
  dplyr::mutate(n_signif = ifelse(sign == "signif_down",
                           -n_signif, n_signif)) %>%
  {
    ggplot(.,aes(y = fct_rev(contrast), x = n_signif, fill = sign)) +
      scale_fill_manual(values = c(
        "signif_down" = "dodgerblue", 
        'signif_up' = "tomato")) +
      geom_bar(stat="identity") +
      theme_bw() +
      coord_cartesian(xlim = c(min(.$n_signif),max(.$n_signif)), 
                      ylim = c(1,nrow(.)/2),
                      clip = "off") +
      theme(plot.margin = unit(c(2, 3, 1, 1), "cm"),
            axis.text = element_text(color = "black")) +  
      annotate("text", 
               x = quantile(.[which(.$n_signif <0),]$n_signif,.25), 
               y = nrow(.)/2+1, 
               label = "Downregulated", vjust = 0) +
      annotate("text", 
               x = quantile(.[which(.$n_signif >0),]$n_signif,.75), 
               y = nrow(.)/2+1, 
               label = "Upregulated", vjust = 0) +
      ylab(NULL) +
      xlab("DEGs per contrast (#)") +
      guides(fill = "none") +
      geom_text(data = ., 
                aes(label = signif_signs,
                    y = contrast, 
                    x = max(n_signif)*1.15, 
                    fill = NULL),
                hjust = 0, size = 3)
}

```

### 3.3.1 - View summary of DESeq2 DEGs

```{r view deseq summary, fig.width = 6, fig.height = 6}
bulk$plots$deseq$pDEGSummaryBars
```

## 3.4 Heatmaps of DE genes (DESeq2)
```{r build DESeq2 heatmap, warning=FALSE}
# prepping list for looping
bulk$plots$deseq$DEheatmaps <- list()
bulk$tables$deseq$scaleGenes <- as.data.frame(t(scale(t(bulk$deseq$normCounts))))

# loop to create list of plots for each contrast
for(i in colnames(bulk$md$contr.matrix)){
  
# defining contrast and pulling out from bulk$md$contr.matrix
  cmSubset <- bulk$md$contr.matrix %>%
    as.data.frame() %>%
    rownames_to_column("group") %>%
    dplyr::select(group, "comparison" = all_of(i))
  
# subsetting gene table
  gene_tbl <- bulk$degTab$deseq %>%
    getTopDegTbl(., contrast_i = i, groupID = "group",
                 arrange_by = "pvalue", direction = "unequal", 
                 padj_cutoff = NULL, slice_n = 50, 
                 filter_pattern = NULL) %>%
    dplyr::select(gene, padj, l2fc = log2FoldChange) %>%
    dplyr::mutate(color = dplyr::case_when(
      is.na(padj) ~ "grey",
      padj >= fdr ~ "black",
      l2fc > 0 ~ "tomato",
      l2fc < 0 ~ "dodgerblue",
      TRUE ~ NA_character_))

# getting gene order for heatmap
  ord <- hclust(dist(bulk$tables$deseq$scaleGenes[c(gene_tbl$gene),]))$order
  
# combining z-score, sample table, and gene table and prepping factors for plotting
 bulk$plots$deseq$DEheatmaps[[i]] <- bulk$tables$deseq$scaleGenes[c(gene_tbl$gene),] %>%
    rownames_to_column("gene") %>%
    tidyr::pivot_longer(!gene,
                 names_to = "sampleID",
                 values_to = "zscore")%>%
    dplyr::inner_join(.,bulk$dge$samples,by="sampleID") %>%
    dplyr::inner_join(.,gene_tbl, by="gene") %>%
    dplyr::inner_join(.,cmSubset, by = "group") %>%
    dplyr::arrange(group,subjectID) %>%    
    dplyr::mutate(
      group = factor(group, levels = unique(group)),
      subjectID = factor(subjectID, levels = unique(subjectID)),
      gene = factor(gene, levels = unique(gene)[ord])) %>%
   {
     ggplot(data = ., 
            aes(x = subjectID, y = gene, fill = zscore)) +
       geom_tile(show.legend = TRUE) +
       scale_fill_gradient2(low="dodgerblue3", mid="white", high="tomato2",
                            limits = c(min(bulk$tables$deseq$scaleGenes),
                                       max(bulk$tables$deseq$scaleGenes))) +
       facet_grid(~ group,              
                  labeller = labeller(group = newLineLabel), 
                  scales="free_x", space="fixed") +
       geom_rect(data = subset(., comparison %in% c(1,-1)), 
                 linewidth = 1.5, fill = NA, colour = "black", 
                 xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
       labs(title = paste0("Top 50 DE Genes in Contrast: ",i), 
            x = NULL, y = NULL) +
       theme_minimal(base_size = 10) +
       theme(plot.title = element_text(size = 16, face = "bold"),
             axis.text.x = element_text(angle = 90, hjust = 1, color = "black"),
             axis.text.y = element_text(color = .$color[ord], size = 10),
             strip.text = element_text(size = 10, face = "bold"))          
   } 
}

```

### 3.4.1 - Plot DE heatmaps (DESeq2)

```{r plot deseq DE heatmaps, fig.width = 10, fig.height = 9}
for(i in colnames(bulk$md$contr.matrix)){
    suppressWarnings(print(ggarrange(bulk$plots$deseq$DEheatmaps[[i]], legend = "bottom")))
}
```

# Part 4 - Compare DEG methods
This chunk compares the DEG l2fc and padj values returned by both methods. There may be other beneficial tests to run

## 4.1 - Generate comparisons
```{r compare DESeq2 and Limma results, warning=FALSE}
# limma results: bulk$degTab$limma
# DESeq2 results: bulk$degTab$deseq

limmaTab <- bulk$degTab$limma %>%
    dplyr::select(contrast, gene, l2fc = log2FoldChange, padj, pvalue) %>%
    dplyr::rename_with(~ paste0("L_", .x, recycle0 = TRUE),
                  all_of(c("l2fc", "padj", "pvalue")))

deseqTab <- bulk$degTab$deseq %>%
    dplyr::select(contrast, gene, l2fc = log2FoldChange, padj, baseMean, pvalue) %>%
    dplyr::rename_with(~ paste0("D_", .x, recycle0 = TRUE),
                  all_of(c("l2fc", "padj", "baseMean", "pvalue")))

fullTab <- full_join(limmaTab,deseqTab, by = c("contrast","gene"))
fullTab <- fullTab %>%
  dplyr::mutate(D_padj = replace_na(D_padj,1)) %>%
  dplyr::mutate(sigCol = ifelse(D_padj < 0.05 & L_padj < 0.05, "red",
                          ifelse(D_padj < 0.05 & L_padj > 0.05, "pink",
                          ifelse(D_padj > 0.05 & L_padj < 0.05, "orange","grey"))))

# plot with all samples
pbase <- ggplot(data = fullTab,
             aes(x = L_l2fc, y = D_l2fc, fill = sigCol), 
             color = "black", shape = 21) +
  theme_bw(base_size = 8) + 
  scale_fill_identity() +
  xlab("limma l2fc") +
  ylab("DESeq2 l2fc") +
  xlim(c(min(fullTab$L_l2fc),max(fullTab$L_l2fc))) +
  ylim(c(min(fullTab$D_l2fc),max(fullTab$D_l2fc))) 

p1 <- pbase + 
  geom_point(data = fullTab[which(fullTab$sigCol == "grey"),],
             aes(x = L_l2fc, y = D_l2fc, fill = sigCol), color = "black", shape = 21) +
  geom_point(data = fullTab[which(fullTab$sigCol != "grey"),],
             aes(x = L_l2fc, y = D_l2fc, fill = sigCol), color = "black", shape = 21) +
  xlab(NULL) +
  scale_fill_identity(guide = "legend", 
                      labels = c("none", "limma only","DESeq2 only","both methods")) 
  
p2 <- pbase + 
  geom_point(data = fullTab[which(fullTab$sigCol == "red"),],
             aes(x = L_l2fc, y = D_l2fc, fill = sigCol), color = "black", shape = 21) +
  geom_smooth(data = fullTab[which(fullTab$sigCol == "red"),],
             aes(x = L_l2fc, y = D_l2fc, color = "black"), linewidth = 0.5, 
             method = "lm", se = FALSE, show.legend = FALSE) +
  xlab(NULL) +
  ylab(NULL) 

p3 <- pbase + 
  geom_point(data = fullTab[which(fullTab$sigCol == "orange"),],
             aes(x = L_l2fc, y = D_l2fc, fill = sigCol), color = "black", shape = 21) +
  geom_smooth(data = fullTab[which(fullTab$sigCol == "orange"),],
             aes(x = L_l2fc, y = D_l2fc, color = "black"), linewidth = 0.5, 
             method = "lm", se = FALSE, show.legend = FALSE) 

p4 <- pbase + 
  geom_point(data = fullTab[which(fullTab$sigCol == "pink"),],
             aes(x = L_l2fc, y = D_l2fc, fill = sigCol), color = "black", shape = 21) +
  geom_smooth(data = fullTab[which(fullTab$sigCol == "pink"),],
             aes(x = L_l2fc, y = D_l2fc, color = "black"), linewidth = 0.5,
             method = "lm", se = FALSE, show.legend = FALSE) +
  ylab(NULL) 


# compare log baseMean of DEGs per method
p5 <- ggplot() +
  geom_boxplot(data = fullTab, aes(x = sigCol, y = log(D_baseMean), fill = sigCol),
               outlier.stroke = 0.2) +
  scale_fill_identity() +
  theme_bw(base_size = 8) +
  labs(title = "Log baseMean") +
  xlab(NULL) +
  scale_x_discrete(labels = c("none", "limma\nonly","DESeq2\nonly","both\nmethods")) +
  ylab("Log baseMean") 

# compare DESeq2 L2FC of DEGs
p6 <- ggplot() +
  geom_boxplot(data = fullTab, aes(x = sigCol, y = D_l2fc, fill = sigCol),
               outlier.stroke = 0.2) +
  scale_fill_identity() +
  theme_bw(base_size = 8) +
  labs(title = "DESeq2 L2FC") +
  xlab(NULL) +
  scale_x_discrete(labels = c("none", "limma\nonly","DESeq2\nonly","both\nmethods")) +
  ylab("DESeq2 L2FC") 

# compare limma L2FC of DEGs
p7 <- ggplot() +
  geom_boxplot(data = fullTab, aes(x = sigCol, y = L_l2fc, fill = sigCol),
               outlier.stroke = 0.2) +
  scale_fill_identity() +
  theme_bw(base_size = 8) +
  labs(title = "limma L2FC") +
  xlab(NULL) +
  scale_x_discrete(labels = c("none", "limma\nonly","DESeq2\nonly","both\nmethods")) +
  ylab("limma L2FC") 
  
bulk$plots$compareMethod1 <- ggarrange(p1,p2,p3,p4, 
                                       ncol = 2, nrow = 2,
                                       common.legend = TRUE)
bulk$plots$compareMethod2 <- ggarrange(p5,p6,p7, 
                                       ncol = 3, nrow = 1,
                                       common.legend = TRUE)
```

## 4.2 - Plot comparison
```{r compare DESeq2 and Limma results, width = 10, height = 10, message = FALSE}
bulk$plots$compareMethod1 
bulk$plots$compareMethod2
```

# Part 5 - Export objects

## 5.1 - Prep for export
```{r prep for export}
## increase report count by 1
bulk$config$reportNumber <- sprintf("%02d", as.numeric(bulk$config$reportNumber) + 1)  

## make subdirectories for `outputs` and `rds`
subdir <- paste0("report-", bulk$config$reportNumber)

if (!dir.exists(here("outputs/", subdir))) {
  dir.create(here("outputs/", subdir))}

if (!dir.exists(here("rds/", subdir))) {
  dir.create(here("rds/", subdir))}
```

## 5.2 - Write `bulk` rds
```{r}
bulk %>%
  write_rds(., here("rds", subdir, paste0("bulk-", getDate(), ".rds")))
```

## 5.3 - Export results table from limma
```{r export limma DEGs, warning=FALSE}
resTab_filename_limma <- paste0(
  bulk$config$project, "-",
  bulk$config$reportNumber, "-",
  "DE_results_limma-", getDate(), ".xlsx")

splitResTab <-  bulk$degTab$limma %>% 
  split(.$contrast) %>%
  lapply(.,function(x)
    {x %>% select(contrast,gene,log2FoldChange,AveExpr,pvalue,padj)})

limmaResTab_export <- c(list("Summary" = bulk$tables$limma$nDEGSummary,
                           "AllResults" = bulk$degTab$limma[,c("contrast","gene","log2FoldChange","AveExpr","pvalue","padj")]),
                        splitResTab)

writexl::write_xlsx(limmaResTab_export, here("outputs", subdir, resTab_filename_limma))
```

## 5.4 - Export DESeq2 results table
```{r export DESeq2 DEGs}
resTab_filename_deseq <- paste0(
  bulk$config$project, "-",
  bulk$config$reportNumber, "-",
  "DE_results_DESeq2-",getDate(), ".xlsx")

deseqResTab_export <- c(list("Summary" = bulk$tables$deseq$nDEGSummary,
                           "AllResults" = bulk$degTab$deseq[,c("contrast","gene","log2FoldChange","baseMean","pvalue","padj")]),
                        bulk$deseq$contrasts)

writexl::write_xlsx(deseqResTab_export ,here("outputs", subdir, resTab_filename_deseq))
```

# 6 - Render report

```{r render, eval=FALSE}
# set report file name
report_filename <- paste0(bulk$config$project, "-", 
                          bulk$config$reportNumber, "-",
                          format(Sys.Date(), "%Y%m%d"))

# HTML
rmarkdown::render(input = here("scripts/02-bulkRNA_DE.format.Rmd"),
                  output_file = paste0(report_filename, "_compare.html"),
                  output_dir = here('reports'),
                  output_format = "rmdformats::robobook",
                  #params = list(method="deseq"))
                 #params = list(method="limma"))
                  params = list(compare=TRUE))

rmarkdown::render(input = here("scripts/02-bulkRNA_DE.format.Rmd"),
                  output_file = paste0(report_filename, "_deseq.html"),
                  output_dir = here('reports'),
                  output_format = "rmdformats::robobook",
                  params = list(method="deseq"))
                 #params = list(method="limma"))
                  #params = list(compare=TRUE))

rmarkdown::render(input = here("scripts/02-bulkRNA_DE.format.Rmd"),
                  output_file = paste0(report_filename, "_limma.html"),
                  output_dir = here('reports'),
                  output_format = "rmdformats::robobook",
                  #params = list(method="deseq"))
                  params = list(method="limma"))
                  #params = list(compare=TRUE))

# PDF
tinytex::install_tinytex() # If it is already installed, will ask in terminal if you should reinstall
rmarkdown::render(input = here("scripts/02-bulkRNA_DE.format.Rmd"),
                  output_file = paste0(report_filename, ".pdf"),
                  output_dir = here('reports'),
                  output_format = "pdf_document",
                  params = list(method="deseq"))
                 # params = list(method="limma"))
                 # params = list(compare=TRUE))
```
