---
title: "02-limmaEdgeR-DE-runfile"
author: "ENPRC Genomics Core"
date:  "`r Sys.Date()`"
output: 
  rmdformats::robobook:
    fig_width: 9
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = TRUE,
                      message = TRUE, 
                      echo=FALSE, 
                      fig.width = 8, 
                      fig.height = 4)
```

# Libraries

```{r}
library(tidyverse)
library(limma)
library(edgeR)
remotes::install_github("yerkes-gencore/gencoreBulk#19")
library(gencoreBulk)
library(here)
library(msigdbr)
library(fgsea)
source(here("R/limmaEdgeR_functions.R")) # will move these to gencoreBulk package
```


# Outline of workflow

1. Define contrasts
  - Keep this separate from the `bulk` object for flexibility
2. DE diagnostics
  - Run `voomLmFit()` or `glmQLFit()`
  - Generate results tables
  - Plot MD
  - Plot p-value distributions
3. DE results
  - Summarize DEG counts in each contrast
  - Export results list to excel file

## Report metadata

These variables help assign filenames to outputs and reports so you can rerun this script with different parameters without mixing up or overwriting files you want to track or compare.

```{r}
project_name <- "p00000_template"
report_number <- "02"
report_author <- "Micah Fletcher"
```


# 1 - Read in bulk object rds

```{r}
bulk <- read_rds(here("rds/report-01-QC/bulk-20240222.rds"))
```


# 2 - Define contrasts

See https://bioconductor.org/packages/release/workflows/vignettes/RNAseq123/inst/doc/designmatrices.html for a detailed tutorial on how to set up design formulas, model matrices and contrast matrices for all of the commonly encountered RNAseq study designs.

The example below is for a repeated measures design from the `p23224_Matthew`.

Assuming we'll be using the `limma::duplicateCorrelation()` method of modeling individual as a random effect:
```{r}
bulk$md$design <- stats::model.matrix(~ 0 + grp.day, data = bulk$md$sampleTable)
colnames(bulk$md$design) <- gsub("grp.day", "", colnames(bulk$md$design))

contr.matrix <- makeContrasts(
  # Within-day group effects
  P14_grp3vsgrp2 = grp3.P14 - grp2.P14,
  D00_grp3vsgrp2 = grp3.D00 - grp2.D00,
  D03_grp3vsgrp2 = grp3.D03 - grp2.D03,
  D07_grp3vsgrp2 = grp3.D07 - grp2.D07,
  D28_grp3vsgrp2 = grp3.D28 - grp2.D28,
  # Within-group-2 day effects
  grp2_D03vsD00 = grp2.D03 - grp2.D00,
  grp2_D07vsD00 = grp2.D07 - grp2.D00,
  grp2_D28vsD00 = grp2.D28 - grp2.D00,
  # Within-group-3 day effects
  grp3_D03vsD00 = grp3.D03 - grp3.D00,
  grp3_D07vsD00 = grp3.D07 - grp3.D00,
  grp3_D28vsD00 = grp3.D28 - grp3.D00,
  # # Average group effects
  # Pre_grp3vsgrp2 = (grp3.P14 + grp3.D00)/2 - (grp2.P14 + grp2.D00)/2,
  # Post_grp3vsgrp2 = (grp3.D03 + grp3.D07 + grp3.D28)/3 - (grp2.D03 + grp2.D07 + grp2.D28)/3,
  # # Change-from-baseline ignoring group
  # Overall_D03vsD00 = (grp3.D03 + grp2.D03)/2 - (grp3.D00 + grp2.D00)/2,
  # Overall_D07vsD00 = (grp3.D07 + grp2.D07)/2 - (grp3.D00 + grp2.D00)/2,
  # Overall_D28vsD00 = (grp3.D28 + grp2.D28)/2 - (grp3.D00 + grp2.D00)/2,
  # Interaction between difference-from-baseline and group
  D03vsD00_grp3vsgrp2 = (grp3.D03 - grp3.D00) - (grp2.D03 - grp2.D00),
  D07vsD00_grp3vsgrp2 = (grp3.D07 - grp3.D00) - (grp2.D07 - grp2.D00),
  D28vsD00_grp3vsgrp2 = (grp3.D28 - grp3.D00) - (grp2.D28 - grp2.D00),
  levels = bulk$md$design
  )
```


# 3 - DE diagnostics

## 3.1 - Run `voomLmFit()` or `glmQLFit()`

```{r}
bulk <- fitVoomLm(bulk, 
                  contr.matrix=contr.matrix, 
                  sample.weights = TRUE, 
                  block = bulk$dge$samples$SubjectID)
```


## 3.2 - Generate results tables

```{r}
resultsTable <- createResTable(bulk$contrasts, contr.matrix)
```

There are two easy ways to access tables split by contrast for different purposes using `for()` or `split()`
```{r}
# for (contr in unique(resultsTable$contrast)) {
#   resultsTable %>%
#     filter(contrast == contr)
# }
```

```{r}
# resultsTable %>%
#   split(., .$contrast)
```


## 3.3 - Plot MD

```{r, fig.width = 8, fig.height = 6}
md_data <- resultsTable %>%
    mutate(signif_dir = ifelse(padj >= 0.05, "NotSig",
                           ifelse(log2FoldChange < 0, "Down", "Up")) %>%
             factor(., levels = c("NotSig", "Up", "Down")),
           signif = padj < 0.05)
  
ggplot() +
  geom_point(data = filter(md_data, !signif),
             aes(x = AveExpr, y = log2FoldChange), 
             color = "black", size = 0.1) +
  geom_point(data = filter(md_data, signif), 
             aes(x = AveExpr, y = log2FoldChange, color = signif_dir), 
             size = 0.1) +
  ylim(-max(abs(md_data$log2FoldChange)), max(abs(md_data$log2FoldChange))) +
  scale_color_manual(values = c("black", "red", "blue"), drop = FALSE) +
  theme_bw() +
  facet_wrap(~contrast)
```


## 3.4 - Plot p-value distributions

```{r, fig.width = 8, fig.height = 6}
resultsTable %>%
      mutate(`pBH < 0.05` = ifelse(padj < 0.05, "TRUE", "FALSE") %>%
               factor(., levels = c("TRUE", "FALSE"))) %>%
    ggplot(data = ., aes(x = pvalue, fill = `pBH < 0.05`)) +
      geom_histogram(breaks = seq(0,1,0.01)) +
      scale_fill_manual(values = c("red", "black"), drop=FALSE) +
      theme_bw() +
    facet_wrap(~contrast, scales = "free_y")
```

# 4 - DE results

## 4.1 - Summarize DEG counts in each contrast

```{r}
n_deg_summary <- resultsTable %>% 
  group_by(contrast) %>%
    dplyr::rename(l2fc = log2FoldChange, pBH = padj) %>%
    summarize(signif_down = sum(pBH < 0.05 & l2fc < 0),
              nonsignif = sum(pBH >= 0.05),
              signif_up = sum(pBH < 0.05 & l2fc > 0)) %>%
  mutate(signif = signif_up + signif_down)
```

### Table
```{r}
n_deg_summary %>%
  mutate(signif_signs = paste0(signif_up, " ↑ / ", signif_down, " ↓")) %>%
  select(contrast, signif_signs) %>%
  knitr::kable()
```

### Bar plots
```{r, fig.width = 12, fig.height = 8}
n_deg_summary %>%
  pivot_longer(cols = c(signif_down, signif_up), 
               names_to = "sign", 
               values_to = "n_signif") %>%
  mutate(n_signif = ifelse(sign == "signif_down",
                           -n_signif, n_signif)) %>%
  ggplot(aes(y = fct_rev(contrast), x = n_signif, fill = sign)) +
  geom_bar(stat="identity") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  theme_bw() +
  ylab("contrast")
```

## 4.2 - Export results list to excel file

## Create a subdir name and keep it consistent across dirs

```{r}
## Subdir name for `outputs` and `rds`
subdir <- paste0("report-", report_number, "-DE")
## Formatted date for filenames
getDate <- function() {
  format(Sys.Date(), "%Y%m%d")
}
```

Create output dir for this report if it doesn't already exist
```{r}
if (!dir.exists(here("outputs/", subdir))) {
  dir.create(here("outputs/", subdir))
}
```

### Export excel file with columns for padj and lfc for each contrast
```{r}
resultsTable_for_export <- unique(resultsTable$contrast) %>%
lapply(., function(contr) {
  resultsTable %>% 
    filter(contrast == contr) %>%
    dplyr::select(gene, l2fc = log2FoldChange, pBH = padj)%>%
    dplyr::rename_with(~ paste0(contr, "_", .x, recycle0 = TRUE),
                  all_of(c("l2fc", "pBH")))
}) %>% 
  reduce(full_join, by = "gene")

resultsTable_filename <- paste0(
  project_name, "-", 
  report_number, "-",
  "DE_results_tables-", 
  getDate(), ".xlsx"
  )

c(list("Summary" = n_deg_summary), 
  list("ResultsTable" = resultsTable_for_export)) %>%
writexl::write_xlsx(here("outputs/", subdir, resultsTable_filename))
```

# 5 - GSEA

## 5.1 - Pre-rank DE stats

```{r}
prepGseaStats2 <- function(marker_table, p_col = "p_val", lfc_col = "avg_log2FC") {
  ranked_genes <- marker_table %>% filter(!is.na(.data[[p_col]]))
  ranked_genes$stat <- ranked_genes[[lfc_col]] * -log10(ranked_genes[[p_col]] + .Machine$double.xmin)
  gsea_stats <- setNames(ranked_genes$stat, nm = rownames(ranked_genes))
  gsea_stats <- sort(gsea_stats)
}

gsea_stats <- resultsTable %>%
  filter(!is.na(pvalue)) %>%
  mutate(rank_stat = log2FoldChange * -log10(pvalue + .Machine$double.xmin)) %>% 
  split(.$contrast) %>%
    lapply(., function(y) {
      y %>% select(gene, rank_stat) %>%
        deframe() %>%
        sort()
    })

```

## 5.2 - Prep gene lists

```{r}
msigH <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gene_symbol, gs_name) %>% unstack()

msigGOBP <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP") %>%
  dplyr::select(gene_symbol, gs_name) %>% unstack()

msigCP <- c(
  msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:BIOCARTA") %>%
    dplyr::select(gene_symbol, gs_name) %>% unstack(),
  msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG") %>%
    dplyr::select(gene_symbol, gs_name) %>% unstack(),
  msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:PID") %>%
    dplyr::select(gene_symbol, gs_name) %>% unstack(),
  msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME") %>%
    dplyr::select(gene_symbol, gs_name) %>% unstack(),
  msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:WIKIPATHWAYS") %>%
    dplyr::select(gene_symbol, gs_name) %>% unstack()
)
```

## 5.3 - Run GSEA
```{r}
gsea_results <- list()

gsea_results$H <- gsea_stats %>%
  lapply(., function(stats_vec) {
    fgsea(pathways = msigH, stats = stats_vec,
          minSize = 15, maxSize = 100) %>% arrange(pval)
  })

gsea_results$GOBP <- gsea_stats %>%
  lapply(., function(stats_vec) {
    fgsea(pathways = msigGOBP, stats = stats_vec,
          minSize = 15, maxSize = 100) %>% arrange(pval)
  })

gsea_results$CP <- gsea_stats %>%
  lapply(., function(stats_vec) {
    fgsea(pathways = msigCP, stats = stats_vec,
          minSize = 15, maxSize = 100) %>% arrange(pval)
  })

```

## 5.4 - Export results table with all contrasts in a wide format as excel file
```{r}
gsea_results %>%
  lapply(., function(x) {
    bind_rows(x, .id = "contrast")
    }) %>%
  bind_rows(.id = "database")


## Format results for excel/reports
gsea_results_fmt <- lapply(names(gsea_results), function(db) {
    lapply(gsea_results[[db]], function(gsea_results_contrast) {
      gsea_results_contrast %>%
        as_tibble() %>%
        arrange(padj) %>%
        rowwise() %>%
        mutate(leading_edge = paste(leadingEdge, collapse = ", ")) %>%
        ungroup() %>%
        dplyr::select(pathway, ES, NES, log2err, pval, padj, leading_edge)
    })
})
names(gsea_results_fmt) <- names(gsea_results)
# Reorder contrast in order of levels to keep consistent order in reports
gsea_results_fmt <- gsea_results_fmt %>% 
  lapply(., function(results_db) {
    results_db <- results_db[levels(resultsTable$contrast)] 
})

## One sheet per database, each all contrasts side-by-side
gseaResults_compact <- lapply(names(gsea_results), function(db) {
    lapply(names(gsea_results[[db]]), function(contrast) {
      gsea_results_fmt[[db]][[contrast]] %>%
        dplyr::select(pathway, NES, pBH = padj) %>%
        dplyr::rename_with(~ paste0(contrast, "_", .x, recycle0 = TRUE), all_of(c("NES", "pBH")))
          }) %>%
    purrr::reduce(full_join, by = "pathway")
})
names(gseaResults_compact) <- names(gsea_results_fmt)

GSEA_filename <- paste0(
  project_name, "-", 
  report_number, "-",
  "GSEA_results_tables", "-",
  getDate(), ".xlsx"
  )

gseaResults_compact %>%
  writexl::write_xlsx(here("outputs", subdir, GSEA_filename))
```

# 6 - Write rds

## Write `bulk` rds
```{r}
if (!dir.exists(here("rds/", subdir))) {
  dir.create(here("rds/", subdir))
}

bulk %>%
  write_rds(., here("rds/", subdir, paste0("bulk-", getDate(), ".rds")))
```


# 7 - Render reports

## Set filenames

```{r}
# report_title_root <- "QC"
report_filename_QC <- paste0(project_name, "-", 
                             report_number, "-",
                             "QC-",
                             format(Sys.Date(), "%Y%m%d"), ".html")

# report_title_root <- "DE"
report_filename_DE <- paste0(project_name, "-", 
                             report_number, "-",
                             "DE-",
                             format(Sys.Date(), "%Y%m%d"), ".html")
```

## Number of genes in heatmaps

```{r}
n_genes <- 30
n_pathways <- 30
```


## Render
```{r}
rmarkdown::render(input = here("scripts/02-limmaEdgeR-DE-format.Rmd"),
                  output_file = report_filename_DE,
                  output_dir = here('reports'))
```
