---
params:
  title: "bulk RNAseq QC"
output: 
  rmdformats::robobook:
    fig_width: 9
    toc_depth: 3
title: "`r params$title`"
author: "EPRC Gencore"
date:  "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE, fig.width = 6, fig.height = 4, dev='png')
```

<style type="text/css">
.book .book-body .page-inner {
  max-width: 1600px;
  margin-left: auto;
  margin-right: auto;
}
</style>

## Study overview

```{r}
sample_metadata_table
```

## QC

### Mapping

Outcomes from mapping reads to reference genome. Reads need to be 'identified' or successfully mapped to a single feature to be used for gene expression analysis.

```{r mapping_plot, fig.width=12, fig.height=8}
analysis$mapping_bins_plot
```

### Relative Log Expression (RLE) normalization

Library size normalization of read counts with hypothesis that most genes are not differentially expressed

```{r rle, fig.width = 12, fig.height=8}
analysis$RLE_raw 
analysis$RLE_norm 
```

### PCA

Exploring projections of expression profiles onto low dimensional space using a limited number of highly variable features. This will give us an estimate of similarity between samples, but will not account for all genes.

```{r pca, fig.width=12, fig.height=8}
print(analysis$pca)

## Plotly can work, but can give some weird legends as multiple legends aren't
## supported. If you have only one factor in the legend you should be okay,
## and the plotly implementation will allow hover info
#ggplotly(analysis$pca, hover = 'text')
```


